Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-12 paper.tex
Ispell

  In defiance of Hinchliffe's rule, this article sets out to
  demonstrate that its title can be answered by the word ``yes''. We
  show that modern symmetric ciphers can be modeled through a small
  set of algebraic structures (Boolean algebra, Naperian and foldable
  functors). We reveal that these ciphers exhibit some interesting
  compositional structure at the type-level, which enables systematic
  code transformation, known in the cryptographical folklore as
  ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
  development providing the specification of two ciphers, Skinny and
  Gift, deriving a bitsliced and fixsliced implementation for each and
  proving their correctness with respect to their specifications.

@@

Following the NSA involvement in the design of the DES diffusion
layer~\citep{johnson:cold-war-crypto}, cryptographers have been
advocating for ``nothing-up-my-sleeve'' designs (which does not make
magic tricks impossible, just
harder~\citep{bernstein:something-up-my-sleeves}). For example, the
diffusion layer of the AES standard is based on the usual
multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state of
modern ciphers is thus equipped with some matricial structure
(oftentimes, manipulated as a 2D or 3D array of bits): the cipher is
not specified over a flat sequence of individual bits but over a
n-dimensional array of bits.

@@

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
programmign language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to suit this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographical primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
correctness and, even more usefully, test our code, this also enables
us to easily prototype a fixslicing compiler, using Coq's
\texttt{autorewrite} tactics to emulate a transformation-based simplification
engine~\citep{peyton-jones:haskell-optim} (part of an imaginary optimizing
compiler).

@@

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
latter worked on the specification, bitslicing and fixslicing of the
Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
ciphers are provided in the aaccompanying source
code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
we shall focus exclusively on Skinny as our running example here.

@@

Operationally, reindexing effects a change of the iteration order of
the composed data container: we go from supporting iteration over
\coqe!G! within an external iteration over \coqe!F! to an interation
over \coqe!F! within an external iteration over \coqe!G!.

@@

Considering (very carefully!) the instanciations of \coqe!subcells_!,
\coqe!add_round_key_! and \coqe!psi_! (which is part of
\coqe!mixcolumns_!), we check that these can be efficiently
implemented over machine words: they only involve boolean operations,
applied bitwise over \coqe!reg32!.

@@

Indeed, aside from the logical operations, the rotations are in fact
almost free thanks to the \textsc{Arm} Cortex barrel
shifters.\footnote{By this remark, we hope to have established our
credentials as budding cryptographers.}

@@

  \coqfrom{Skinny/FixsliceProof}{tau_phi4}
  %
  where \coqe!tau_! is defined as
  %
  \coqfrom{Skinny/Base}{tau}
  %
  and can be implemented reasonnably efficiently.
\end{prop}

@@

Given 4 (suitably generated) round subkeys, we merely have to chain 4
rounds with the corresponding diffusion layer, ending with \coqe!tau_!
to resynchronize back to identity
%
\coqfrom{Skinny/Fixslice}{round}

@@

The correctness statemement is unsurprising, modulo some light
bureaucracy to ensure that round subkeys have been suitably generated
%
\begin{theo}
  \coqfrom{Skinny/FixsliceProof}{theorem}
\end{theo}

>>>>

  In defiance of Hinchliffe's rule, this article sets out to
  demonstrate that its title can be answered by the word ``yes''. We
  show that modern symmetric ciphers can be modeled through a small
  set of algebraic structures (Boolean algebra, Naperian and foldable
  functors). We reveal that these ciphers exhibit some interesting
  compositional structure at the type-level, which enables systematic
  code transformation, known in the cryptographic folklore as
  ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
  development providing the specification of two ciphers, Skinny and
  Gift, deriving a bitsliced and fixsliced implementation for each and
  proving their correctness with respect to their specifications.

@@

Following the NSA involvement in the design of the DES diffusion
layer~\citep{johnson:cold-war-crypto}, cryptographers have been
advocating for ``nothing-up-my-sleeve'' designs (which does not make
magic tricks impossible, just
harder~\citep{bernstein:something-up-my-sleeves}). For example, the
diffusion layer of the AES standard is based on the usual
multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state of
modern ciphers operate over a certain matrix (oftentimes, a 2D or 3D
array of bits): the cipher is not specified over a flat sequence of
individual bits but over a n-dimensional array of bits.

@@

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
programming language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to suit this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographic primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
correctness and, even more usefully, test our code, this also enables
us to easily prototype a fixslicing compiler, using Coq's
\texttt{autorewrite} tactics to emulate a transformation-based simplification
engine~\citep{peyton-jones:haskell-optim} (part of an imaginary optimizing
compiler).

@@

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
latter worked on the specification, bitslicing and fixslicing of the
Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
ciphers are provided in the accompanying source
code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
we shall focus exclusively on Skinny as our running example here.

@@

Operationally, reindexing effects a change of the iteration order of
the composed data container: we go from supporting iteration over
\coqe!G! within an external iteration over \coqe!F! to an iteration
over \coqe!F! within an external iteration over \coqe!G!.

@@

Considering (very carefully!) the instances of \coqe!subcells_!,
\coqe!add_round_key_! and \coqe!psi_! (which is part of
\coqe!mixcolumns_!), we check that these can be efficiently
implemented over machine words: they only involve boolean operations,
applied bit-wise over \coqe!reg32!.

@@

Indeed, aside from the logical operations, the rotations are in fact
almost free thanks to the \textsc{Arm} Cortex barrel
shifter.\footnote{By this remark, we hope to have established our
credentials as budding cryptographers.}

@@

  \coqfrom{Skinny/FixsliceProof}{tau_phi4}
  %
  where \coqe!tau_! is defined as
  %
  \coqfrom{Skinny/Base}{tau}
  %
  and can be implemented reasonably efficiently.
\end{prop}

@@

Given 4 (suitably generated) round subkeys, we merely have to chain 4
rounds with the corresponding diffusion layer, ending with \coqe!tau_!
to re-synchronize back to identity
%
\coqfrom{Skinny/Fixslice}{round}

@@

The correctness statement is unsurprising, modulo some light
bureaucracy to ensure that round subkeys have been suitably generated
%
\begin{theo}
  \coqfrom{Skinny/FixsliceProof}{theorem}
\end{theo}
